The Galactic Midplane Is Not a Plane: Implications for Dynamical Analysis with {\em Gaia} Data and Beyond

### 2019-05-03 paper/ms.tex
Make changes to abstract

Stellar actions, computed for stars from their six-dimensional phase space
measurements and an assumed Galactic potential, are used to label and
distinguish orbits. In principle, actions have the attractive quality of being
invariant and thus good labels of a star's orbit. Such analyses are often used
to interpret data from the {\em Gaia} mission. However, inaccurate
measurements of the phase space position and uncertainties in the assumed
Galactic potential induce an error in the computed action. We show that, in
addition to these complications, a systematic bias in the phase space position
induces a phase-dependence in the actions, which we interpret as a systematic
error. The induced error distribution is non-Gaussian and bimodal, with
neither of the modes peaking on the null value. A variation in the vertical
position midplane of $\sim15\,\pc$ for a thin disk orbit or $\sim120\,\pc$ for
a thick disk orbit is sufficient to induce a $25\,\%$ systematic error in the
vertical action $J_z$ computed under the assumption of an axisymmetric
potential. Furthermore, we show that the local midplane varies by
$\sim100\,\pc$ at the Solar circle in Milky Way-mass cosmological zoom-in
simulations from the FIRE collaboration. Thus, current state-of-the-art action
calculations --- which assume the global and local midplanes are the same ---
include a systematic vertical offset variation. Variation in the local
standard of rest induces similar issues. The variation of the midplane must be
taken into account when performing dynamical modelling across the large
regions of the disk accessible to {\em Gaia}.

>>>>

Stellar actions, computed for stars from their six-dimensional phase space
measurements and an assumed Galactic potential, are used to label and
distinguish orbits. In principle, actions have the attractive quality of being
invariant and thus good labels of a star's orbit. Such analyses are often used
to interpret data from the {\em Gaia} mission. However, inaccurate
measurements of the phase space position and uncertainties in the assumed
Galactic potential induce an error in the computed action. We show that, in
addition to these complications, a systematic bias in the phase space position
induces a phase-dependence in the actions, which we interpret as a systematic
error. An offset in the vertical
position of $\sim15\,\pc$ for a thin disk orbit or $\sim120\,\pc$ for
a thick disk orbit is sufficient to induce a $25\,\%$ systematic error in the
vertical action $J_z$ computed under the assumption of an axisymmetric
potential. The induced error distribution is non-Gaussian and bimodal, with
neither of the modes peaking on the null value. Furthermore, we show that the local midplane varies by
$\sim100\,\pc$ at the Solar circle in Milky Way-mass cosmological zoom-in
simulations from the FIRE project. From observations that the mean vertical velocity of the Milky Way disk varies by $\sim5\,\kms$ with radius, we estimate that the true midplane variations are of order $\sim100\,\pc$. Thus, current state-of-the-art action
calculations --- which assume the global and local midplanes are the same ---
are likely to include a systematic vertical offset variation depending on the volume considered. Variation in the local
standard of rest induces similar issues. The variation of the midplane must be
taken into account when performing dynamical modelling across the large
regions of the disk accessible to {\em Gaia}.
